---
title: "Report"
author: "Qi Qi"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: pdf_document
---

```{r setup, include=FALSE}
library("dplyr")
library("ggpubr")
library("goft")
library("mclust")
library("mixtools")
library("mixR")
knitr::opts_chunk$set(echo = TRUE)
```


## Distribution of Experimental Data
Test the normality of experimental data set:

```{r}
test <- read.csv("Data for Root Cause Determination - Test Data.csv")
dat <- read.csv("Data for Root Cause Determination.csv")
mean(test$Response[test$Group=="Test Group - Root Cause"])
sd(test$Response[test$Group=="Test Group - Root Cause"])
mean(test$Response[test$Group=="Test Group - No Root Cause"])
sd(test$Response[test$Group=="Test Group - No Root Cause"])

ggqqplot(test[test$Group == "Test Group - Root Cause",]$Response)
shapiro.test(test[test$Group == "Test Group - Root Cause",]$Response)
ggqqplot(test[test$Group == "Test Group - No Root Cause",]$Response)
shapiro.test(test[test$Group == "Test Group - No Root Cause",]$Response)

```

From above result, we know for each group the data is normally distributed. Since sample size is 30 in each group, we do not have much power. Then I also test goodness of fit of Gamma distribution:


```{r}
gamma_test(test[test$Group == "Test Group - Root Cause",]$Response)
gamma_test(test[test$Group == "Test Group - No Root Cause",]$Response)
```
Above results also show gamma distribution fits the data set.

Then I propose mixed normal distribution and mixed gamma distribution of original data set.

## Mixture of Normal Distributions

Let $z_i$ be the index of the Gaussian distribution from which $x_i$ is sampled. The parameters to be estimated is $(\mu_1, \mu_2,\sigma_1^2, \sigma_2^2,\xi_1,\xi_2)$, where $\xi_1+\xi_2=1$. Let $\theta_t=(\mu_{t1}, \mu_{t2},\sigma_{t1}^2, \sigma_{t2}^2,\xi_{t1},\xi_{t2})$
$$Q(\theta|\theta_t)=\sum_zp(z|x,\theta_t)\ln p(x,z|\theta)=\sum_{i=1}^n\sum_{k=1}^2p(z_i=k|x_i,\theta_t)\ln p(x_i,z_i=k|\theta)$$
Let $w_{ik}=p(z_i=k|x_i,\theta_t)$, then
$$w_{ik}=\frac{p(z_i=k,x_i,|\theta_t)}{\sum_{k=1}^2p(z_i=k,x_i|\theta_t)}=\frac{\xi_{tk}\phi(x_i|\mu_{tk},\sigma_{tk}^2)}{\sum_{k=1}^2\xi_{tk}\phi(x_i|\mu_{tk},\sigma_{tk}^2)}$$
$$Q(\theta|\theta_t)=\sum_{k=1}^2\sum_{i=1}^nw_{ik}\ln (\xi_k/\sqrt{2\pi})-\frac{1}{2}\sum_{k=1}^2\sum_{i=1}^nw_{ik}\ln \sigma_k^2-\frac{1}{2}\sum_{k=1}^2\sum_{i=1}^nw_{ik}\frac{(x_i-\mu_k)^2}{\sigma_k^2}$$
$$\frac{\partial Q(\theta|\theta_t)}{\partial \mu_k}=0 \Rightarrow \mu_k=\frac{\sum_{i=1}^nw_{ik}x_i}{\sum_{i=1}^nw_{ik}}$$
$$\frac{\partial Q(\theta|\theta_t)}{\partial \sigma_k^2}=0 \Rightarrow \sigma_k^2=\frac{\sum_{i=1}^nw_{ik}(x_i-\mu_k)^2}{\sum_{i=1}^nw_{ik}}$$
$$\xi_k=\frac{1}{n}\sum_{i=1}^nw_{ik}$$

### Implement in R

```{r}
mynormalmixEM <- function(x, xi1, xi2, mu1, mu2, sigma1, sigma2, maxit, tol){
  n <- length(x)
  w1 <- double(n)
  w2 <- double(n)
  for (i in 1:maxit){
     for (j in 1:n){
      w1[j] <- xi1 * dnorm(x[j], mu1, sigma1) / (xi1 * dnorm(x[j], mu1, sigma1) + xi2 * dnorm(x[j], mu2, sigma2))
      w2[j] <- 1 - w1[j]
    }
    xi1new <- mean(w1)
    xi2new <- mean(w2)
    mu1new <- sum(w1 * x) / sum(w1)
    mu2new <- sum(w2 * x) / sum(w2)
    sigma1new <- sqrt(sum(w1 * (x - mu1new)^2) / sum(w1))
    sigma2new <- sqrt(sum(w2 * (x - mu2new)^2) / sum(w2))
    if (max(abs(xi1new - xi1), abs(xi2new - xi2), abs(mu1new - mu1), abs(mu2new - mu2), abs(sigma1new - sigma1), abs(sigma2new- sigma2)) < tol){
      xi <- c(xi1new, xi2new)
      mu <- c(mu1new, mu2new)
      sigma <- c(sigma1new, sigma2new)
      iter <- i
      return(list(xi, mu, sigma, iter))
    }
    xi1 <- xi1new
    xi2 <- xi2new
    mu1 <- mu1new
    mu2 <- mu2new
    sigma1 <- sigma1new
    sigma2 <- sigma2new
  }
}

mixnormal <- function(n, xi, mu1, mu2, sigma1, sigma2){
  x <- double(n)
  for (i in 1:n){
    u <- runif(1)
    x[i] <- rnorm(1, ifelse(u < xi, mu1, mu2), ifelse(u < xi, sigma1, sigma2))
  }
  x
}
## simulated data
data <- mixnormal(1000, .6, 3, 8, 1, 1)
mynormalmixEM(data, .5, .5, 6, 7, 1, 4, 1e5, 1e-5)  
out.1 <- normalmixEM(data, arbvar = FALSE, epsilon = 1e-03, fast=TRUE)
summary(out.1)
fit.1 <- mixfit(data, ncomp = 2, family = "normal")
fit.1


## real data
mynormalmixEM(test$Response, .5, .5, 100, 110, 3, 5, 1e5, 1e-5)  
mynormalmixEM(dat$Response, .5, .5, 100, 110, 3, 5, 1e5, 1e-5) 

## Using "mixtools" package
out.1 <- normalmixEM(test$Response, arbvar = FALSE, epsilon = 1e-03, fast=TRUE)
summary(out.1)
plot(out.1, density = TRUE, w = 1.1)

out.2 <- normalmixEM(dat$Response, arbvar = FALSE, epsilon = 1e-03, fast=TRUE)
summary(out.2)
plot(out.2, density = TRUE, w = 1.1)

## Using "mixR" package
fit.1 <- mixfit(test$Response, ncomp = 2, family = "normal")
fit.1
plot(fit.1)

fit.2 <- mixfit(dat$Response, ncomp = 2, family = "normal")
fit.2
plot(fit.2)
```

## Mixture of Gamma Distributions

Let $z_i$ be the index of the Gaussian distribution from which $x_i$ is sampled. The parameters to be estimated is $(\alpha_1, \alpha_2,\beta_1, \beta_2,\xi_1,\xi_2)$, where $\xi_1+\xi_2=1$. Let $\theta_t=(\alpha_{t1}, \alpha_{t2},\beta_{t1}, \beta_{t2},\xi_{t1},\xi_{t2})$
$$Q(\theta|\theta_t)=\sum_zp(z|x,\theta_t)\ln p(x,z|\theta)=\sum_{i=1}^n\sum_{k=1}^2p(z_i=k|x_i,\theta_t)\ln p(x_i,z_i=k|\theta)$$
Let $w_{ik}=p(z_i=k|x_i,\theta_t)$, then
$$w_{ik}=\frac{p(z_i=k,x_i,|\theta_t)}{\sum_{k=1}^2p(z_i=k,x_i|\theta_t)}=\frac{\xi_{tk}\Gamma(x_i|\alpha_{tk},\beta_{tk})}{\sum_{k=1}^2\xi_{tk}\Gamma(x_i|\alpha_{tk},\beta_{tk})}$$
$$Q(\theta|\theta_t)=\sum_{k=1}^2\sum_{i=1}^nw_{ik}\ln \xi_k-\sum_{k=1}^2\sum_{i=1}^nw_{ik}\ln \Gamma(\alpha_k)-\sum_{k=1}^2\sum_{i=1}^nw_{ik}\alpha_k \ln \beta_k +\sum_{k=1}^2\sum_{i=1}^nw_{ik}(\alpha_k-1)\ln x_i-\sum_{k=1}^2\sum_{i=1}^n\frac{w_{ik}x_i}{\beta_k}$$
$$\frac{\partial Q(\theta|\theta_t)}{\partial\beta_k}=0 \Rightarrow \beta_k=\frac{\sum_{i=1}^nw_{ik}x_i}{\alpha_k\sum_{i=1}^nw_{ik}}$$
$$\xi_k=\frac{1}{n}\sum_{i=1}^nw_{ik}$$

### Implement in R

```{r}
## Using "mixtools" package
out.1 <- gammamixEM(test$Response)
out.1$lambda
out.1$gamma.pars


out.2 <- gammamixEM(dat$Response)
out.2$lambda
out.2$gamma.pars

## Using "mixR" package
fit.1 <- mixfit(test$Response, ncomp = 2, family = "gamma")
fit.1
plot(fit.1)

fit.2 <- mixfit(dat$Response, ncomp = 2, family = "gamma")
fit.2
plot(fit.2)
```